---
title: "Project: COVID-19 Maps"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    theme: flatly
---

```{r setup, include=FALSE, message=FALSE}
library(sf)
library(tmap)
library(tidyverse)
library(rgeoda)
library(biscale)
library(cowplot)
library(RColorBrewer)
library(reticulate)
```

```{r, include=FALSE, warning=FALSE, message=FALSE}
# Read-in shp
covid_zip <- st_read("/Users/kuiyan/Desktop/R Projects/kirankui.github.io/data/NYC_covid19_zip.shp")

```


> *“📅 2021 was a critical year in the pandemic — vaccines were available 💉, but the burden of COVID-19 remained deeply unequal ⚠️. In a city as large, diverse, and densely populated as New York City 🗽, analyzing spatial data helps reveal how geography and mortality intersected 💵💀. This kind of spatial analysis 📊 provides vital lessons for designing smarter, more equitable public health responses 🏥 in future crises.”*

---

## 💵💀🗽 Poverty & COVID-19 Deaths in NYC (2020–2021)

📊 This section explores how **poverty levels** across New York City’s boroughs 🏙️ may have influenced **COVID-19 mortality** 💀 during the first two years of the pandemic 🦠.  
By examining this data, we aim to uncover patterns 🔍 that can inform smarter public health planning 🧠 and resource allocation 🏥 in future emergencies 🚨.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tm_shape(covid_zip) +
  tm_polygons(col =  "per_pov",
              style = "quantile",
              n = 4,
              title = "Percent Poor",
              palette = "Purples",
              border.col = "black") +
  tm_shape(covid_zip) +
  tm_dots(col = "black",
          size = "r_death",
          style = "quantile",
          title = c("Death Rate"),
          palette = "Purples",
          border.col = "black")

```

```{r, include=FALSE, warning=FALSE, message=FALSE}
deaths <- st_read("/Users/kuiyan/Desktop/R Projects/kirankui.github.io/data/NYC_covid19_zip.shp")

clusters <-  st_read("/Users/kuiyan/Desktop/R Projects/kirankui.github.io/data/clusters_death.shp")

sites <- st_read("/Users/kuiyan/Desktop/R Projects/kirankui.github.io/data/NYC_covid19_vacc_sites.shp")

deaths = deaths %>% 
  rename("percent_over_65_yo" = per_pop65,
         "percent_high_school" = per_highed,
         "percent_poverty" = per_pov,
         "percent_income" = per_pubinc,
         "infection" =  r_covid) 
  
```

---

## 🗺️💀📊 Interactive Map: COVID-19 Deaths in NYC (2020–2021)

🖱️ *Click on neighborhoods 🏘️ to explore sociodemographic data 📂 — including poverty rates 💵, median income 💰, population density 🧍‍♂️🧍‍♀️, and COVID-19 death counts ⚰️.*

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tmap_mode("view")

tm_shape(deaths) +
  tm_polygons(col = "r_death",
              palette = "GnBu",
              alpha = .8,
              style = "quantile",
              n = 4,
              border.col = "white",
              lwd = .5,
              title = "Death Rate",
              interactive = TRUE,
              popup.vars = c("percent_over_65_yo",  "percent_high_school", "percent_poverty", "infection"))

```

---

## ✅ Conclusion

This analysis reveals a clear pattern: neighborhoods with higher poverty rates 💵 experienced greater COVID-19 mortality 💀 in New York City 🗽 during 2020–2021.

Despite the availability of vaccines in 2021 💉, the burden of the pandemic remained deeply unequal ⚠️ — especially in communities already facing structural disadvantages. These findings underscore how socioeconomic conditions directly shape public health outcomes 🧠.

📊 By analyzing neighborhood-level data, we gain critical insights into where health systems and emergency response efforts must improve — not just for future pandemics 🦠, but for everyday city planning 🏥.

The data speaks clearly: **place, income, and survival are deeply connected**.
