---
title: "Project: Hot Summer"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    theme: flatly
---

```{r setup, include=FALSE}

library(tidyverse)
library(dplyr)
library(readxl)
library(ggplot2)
library(openair)
library(lubridate)
library(plotly)
```

> *“Boston’s 2023 summer saw record heat waves 🌡️, pushing the city to declare heat emergencies 🏘. Understanding where and how heat can occur has become increasingly important for shaping policy, strengthening community resilience, and ensuring climate preparedness across cities like Boston 🧭🌆. While agencies like MassDEP already collect extensive environmental monitoring data, the next step is to understand what it means—analyzing heat trends to help identify health risks ⚠️, guide targeted interventions, and inform city planning 📊🔍.”*

---

## 📍 Project Introduction and Context

This project explores environmental data collected from **Reggie Wong Park** in Boston during **July and August 2023**. The goal is to uncover patterns and relationships between:

- 🌡️ **Outdoor Temperature**  
- 💧 **Relative Humidity**  
- 🌬 **Other weather conditions (e.g., wind speed)**

---

## 🔍 Methodology: End-to-End Data Pipeline & Time Series Analysis

Weather data was collected every **10 minutes** using a **Kestrel sensor**. I built an end-to-end data pipeline in R for **data cleaning (QC)** and applied **time series analysis** to examine how features like temperature and humidity influence local environmental conditions.

---

## 📈 What is Time Series Analysis?

This method is widely applicable across industries:

- **Healthcare** 🏥: Patient admissions, disease trends, treatment effectiveness  
- **Retail** 🛒: Sales patterns, customer foot traffic, inventory flow  
- **Finance** 💹: Stock prices, interest rates, cryptocurrency trends  
- **Manufacturing** 🏭: Output tracking, supply chain delays, equipment maintenance

---

## 📂 Dataset Source

Data were collected using **Kestrel Sensors** at Reggie Wong Park on **August 2, 2023**. Please email me for more information about this dataset.

---

## 💻 Programming & Tools Used

- **R**: `tidyverse`, `dplyr`, `ggplot2`, `plotly`, `lubridate`, `openair`  
- **Version control**: Git

---

```{r, echo=FALSE}

kestrel_df <- read_xlsx("./data/Reggie_Kestrel_08022023.xlsx", sheet=1, range = "A5:G6228") %>% 
  janitor::clean_names() %>% 
  slice(-1) %>% 
  mutate(temperature = as.numeric(temperature),  
         wet_bulb_temp = as.numeric(wet_bulb_temp), 
         relative_humidity = as.numeric(relative_humidity), 
         station_pressure = as.numeric(station_pressure), 
         heat_index = as.numeric(heat_index), 
         dew_point = as.numeric(dew_point)) %>% 
  rename(date = formatted_date_time)

head(kestrel_df)

```

## 📊 Exploratory Data Analysis

The dataframe `kestrel_df` contains `r ncol(kestrel_df)` variables and `r nrow(kestrel_df)` observations 📁🔢.

This dataset captures a range of environmental metrics 🌡️💧🌬️ collected by the **Kestrel sensor** between **July 8, 2023, at 8:02 PM** and **August 2, 2023, at 3:30 PM** 📆🕒.

📌 Initially, measurements were recorded at **2-minute intervals** ⏱️ until **July 12, 2023, at 7:30 PM**, after which the recording frequency changed to **10-minute intervals** ⏲️ for the remainder of the monitoring period.

---

## 📈 Time Series Analysis

In this section, I apply time series analysis to examine trends and patterns in environmental variables over time 🕒🌡️💧. By analyzing changes in outdoor temperature, relative humidity, and other weather metrics, we can better understand how these factors interact and evolve throughout the monitoring period 🔍📊.

Techniques include:
- 📅 Temporal trend visualization  
- 🔁 Lag analysis  
- 📉 Smoothing and decomposition  
- 📐 Correlation between variables over time

This analysis helps reveal underlying patterns and supports further modeling or prediction efforts 🔮.


```{r, echo=FALSE}
# Date to year, month, day and hour
kestrel_df$date <- ymd_hms(kestrel_df$date)

kestrel_df$year <- year(kestrel_df$date)
kestrel_df$month <- month(kestrel_df$date)
kestrel_df$day <- day(kestrel_df$date)
kestrel_df$hour <- hour(kestrel_df$date)

kestrel_df %>% 
  select(date, year, month, day, hour, everything())

```
📉 I separated the date column into year, month, day and hour.

```{r, echo=FALSE}
# Converted to 10 minute intervals
kestrel_df$date <- floor_date(kestrel_df$date, unit = "10 minutes")

head(kestrel_df)
```
📉 I found the average 10-min values for temperature, humidity…

```{r, echo=FALSE}
# Converted to 1-day intervals
kestrel_daily_df <- timeAverage(kestrel_df, avg.time = "day")

head(kestrel_daily_df)
```
📉 I found the average daily values for temperature, humidity…

```{r, echo=FALSE}
# Converted to monthly intervals
kestrel_monthly_df <- timeAverage(kestrel_df, avg.time = "month")

head(kestrel_monthly_df)
```
📉 I found the average monthly values for temperature, humidity…

```{r, include=FALSE}
# Generate the day of the week (1 = Sunday, 7 = Saturday)
kestrel_df$day_of_week <- wday(kestrel_df$date)

# Generate the day of the week as a factor (e.g., Sunday, Monday)
kestrel_df$day_of_week_name <- wday(kestrel_df$date, label = TRUE, abbr = FALSE)

kestrel_df$day_of_week_name <- factor(kestrel_df$day_of_week_name, 
                                       levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                                  "Thursday", "Friday", "Saturday"))


```

---

## 🧪 Data Methodology Summary

- 📆 Separated the `date` column into **year**, **month**, **day**, and **hour** components for easier time-based grouping.  
- 🔟 Calculated **average 10-minute values** for key variables such as temperature 🌡️ and humidity 💧.  
- 📅 Computed **average daily values** to observe day-to-day trends.  
- 📆 Aggregated **average monthly values** to explore broader seasonal patterns.

📌 **Note:** Only two days of data were recorded in August—**08/01/2023** and **08/02/2023**.

---

## 📊 Visualizations of Trends

This section showcases key visual insights from the dataset using plots to highlight temporal patterns and changes in environmental conditions over time 🕒🌡️💧.

---

### 📆 Comparing July and August Temperatures

Here, we compare **average temperatures** recorded in **July** and the first two days of **August 2023** 🔥🧊. This helps identify short-term seasonal shifts and examine how extreme heat events varied between the two months.

```{r, echo=FALSE}
kestrel_temp <- plot_ly(kestrel_daily_df, 
                        x = ~date, 
                        y = ~temperature, 
                        name = 'Daily Average Temperature (°F)', 
                        type = 'scatter', 
                        mode = 'lines+markers') %>%
  layout(
    title = 'Daily Temperature at Reggie Wong Park in July-Aug 2023',
    yaxis = list(title = 'Temperature (°F)'),
    xaxis = list(title = 'Date'),
    legend = list(
      orientation = 'h',
      y = -0.3,
      x = 0.5,
      xanchor = 'center',
      yanchor = 'bottom'
    )
  )


# View the plot
kestrel_temp


```

---

## 🔎 Findings

- 🌡️ **July 2023** recorded **higher overall temperatures** compared to early August.  
- 📊 We also have **more complete data** available for July.  
- 🎯 Based on this, the analysis will focus primarily on **July 2023** for deeper insights.

---

### 🌦️ Weather Conditions Across Time in July

This section explores how key environmental metrics—such as **temperature**, **humidity**, and other weather conditions—varied throughout **July 2023** 📅📈. By visualizing these changes over time, we can better understand daily patterns, extreme fluctuations, and overall climate trends during the monitoring period.

```{r, echo=FALSE}
timePlot(selectByDate(kestrel_df, year = 2023, month = 7),
pollutant = c("temperature", "relative_humidity"),
y.relation = "free")

```

---

### 🛠️ Data Methodology

I created a **time series plot** to visualize variations in **temperature 🌡️** and **humidity 💧** between **July 8, 2023, at 8:02 PM** and **July 31, 2023, at 11:50 PM**.

---

### 🔍 Findings

- 🔥 **July 13** and **July 24** recorded the **highest temperatures**, reaching nearly **100°F**!  
- 💧 There is an **inverse relationship** between **outdoor temperature** and **humidity**—as temperatures rose, humidity levels tended to drop.

---

### 📆 Weather Conditions Across Days of the Week

This section examines how **temperature** 🌡️ and **humidity** 💧 varied by **day of the week** during July 2023. Identifying patterns across weekdays and weekends can help reveal behavioral or environmental trends, such as differences in urban activity, cooling patterns, or exposure risks.


### Weather Conditions across days of the week
```{r, echo=FALSE}

# temperature
timeVariation(subset(kestrel_df),pollutant = "temperature", ylab = "Temperature (F)")

# Relative Humidity
timeVariation(subset(kestrel_df),pollutant = "relative_humidity", ylab = "Relative Humidity")

```



### 🛠️ Data Methodology

I created a **time series plot** to visualize variations in **temperature 🌡️** and **humidity 💧** across different **days of the week** during **July 2023**.

### 🔍 Findings

- 🌧️ **Monday** showed the **highest humidity** and **lowest temperatures** overall.  
- 🔥 **Wednesday** recorded the **highest temperatures** and **lowest humidity**, suggesting sharper heat and drier conditions midweek.

---

## 🤖 Predictive Models (for July)

In this section, I apply **predictive modeling** techniques to explore how well we can estimate or forecast one weather variable (e.g., temperature 🌡️) based on others like humidity 💧, time of day 🕒, or day of the week 📅.

Using the **July 2023 dataset**, I build models to:

- 🔍 Identify key predictors of outdoor temperature  
- 📈 Quantify relationships between variables  
- 🔮 Explore the potential for short-term forecasting

These models help us better understand patterns in the data and provide a foundation for future applications in environmental monitoring, urban planning, and health response 🚦🏙️.

```{r , include=FALSE, warning=FALSE}
kestrel_july_df = kestrel_df %>% 
  filter(month==7)

fit <- lm(temperature ~ relative_humidity * day, data = kestrel_july_df)
summary(fit)

#base <- lm(temperature ~ relative_humidity + day, data = kestrel_july_df)

#anova(base, fit)

kestrel_july_df$fit_values <- fitted(fit)
```


```{r, echo=FALSE, warning=FALSE}
# Plot the original data points and the fitted line
ggplot(kestrel_july_df) +
  aes(x = relative_humidity, y = temperature) +  
  geom_point(aes(color = factor(day)), alpha = 0.6) +  
  geom_line(aes(x = relative_humidity, y = fit_values, color = factor(day)), size = 1) +  
  theme_minimal() +  
  theme(
    panel.background = element_rect(fill = "white"),  
    plot.background = element_rect(fill = "white"),   
    legend.background = element_rect(fill = "white")  
  ) +
  labs(
    x = "Relative Humidity (%)",                     
    y = "Temperature (°F)",     
    color = "Day in July",                    
    title = "Temperature vs. Relative Humidity by Day in July")


```


```{r, include=FALSE, warning=FALSE}
fit <- lm(temperature ~ relative_humidity * day_of_week_name, data = kestrel_july_df)

summary(fit)

#base <- lm(temperature ~ relative_humidity + day_of_week_name, data = kestrel_july_df)

#anova(base, fit)

kestrel_july_df$fit_values <- fitted(fit)
```

```{r, echo=FALSE, warning=FALSE}
# Plot the original data points and the fitted line
ggplot(kestrel_july_df) +
  aes(x = relative_humidity, y = temperature) +  
  geom_point(aes(color = factor(day_of_week_name)), alpha = 0.6) +  
  geom_line(aes(x = relative_humidity, y = fit_values, color = factor(day_of_week_name)), size = 1) +  
  theme_minimal() +  
  theme(
    panel.background = element_rect(fill = "white"),  
    plot.background = element_rect(fill = "white"),   
    legend.background = element_rect(fill = "white")  
  ) +
  scale_color_brewer(palette = "Set1") + 
  labs(
    x = "Relative Humidity (%)",                     
    y = "Temperature (°F)",     
    color = "Day of the week",                    
    title = "Temperature vs. Relative Humidity by Day of Week in July"  # Added title
  )

```


## 📝 Data Summary and Recommendations for Park Users

Based on weather data collected from **July to August 2023** at Reggie Wong Park:

- 🔥 **July is a hot month**, with temperatures approaching **100°F** during peak periods.  
- 🌡️ **Temperatures fluctuate** throughout the month and begin to **cool slightly** toward the end of July.  
- 📅 **Wednesdays** tend to have the **highest temperatures** and **lowest humidity**, making them especially risky for heat-related stress.

✅ **Recommendation**: Park users should stay hydrated, avoid peak midday hours, and take extra precautions—especially on **Wednesdays**—when exercising or spending extended time outdoors.



